---
title: "Top Spotify Tracks of 2017"
author: "James Le"
output: 
 html_notebook:
  toc: true
  code_folding: hide
---

# Introduction
In this notebook, I'll take a look at the audio features of the tracks in **Spotify's Top Songs of 2017** playlist and try to highlight the common patterns behind the audio features of these songs. 

# Data Processing
First off, let’s import the data and get an idea of what I'm working with.

```{r read, echo=TRUE, message=FALSE, warning=FALSE}
library(readr)
library(plyr)
library(dplyr)
library(ggplot2)
library(formattable)

audio_features <- read.csv('Spotify-Audio-Features.csv')
```

Let's take a look at the summary of this dataset.
```{r summary, echo=TRUE}
summary(audio_features)
```

For the simplicity, I'll convert `duration_ms` variable in seconds rather than miliseconds.
```{r, echo=TRUE, message = FALSE}
audio_features$duration_ms <- round(audio_features$duration_ms / 1000)
colnames(audio_features)[15] <- "duration"
```

# Data Analysis

## Artists dominating the Top List

Let's determine the artists who have more than one song on the Top 100 Songs List.
```{r, echo=TRUE, message = FALSE}
top_artists <- audio_features %>%
    group_by(artists)  %>%
    summarise(number_apperance = n()) %>%
    filter(number_apperance > 1) %>%
    arrange(desc(number_apperance))

top_artists$artists <- factor(top_artists$artists, levels = top_artists$artists[order(top_artists$number_apperance)]) # visualise the list in descending order 

ggplot(top_artists, aes(x = artists, y = number_apperance)) +
    geom_bar(stat = "identity",  fill = "tomato2", width = 0.6 ) + 
    labs(title = "2017 Top Spotify Artists", x = "Artists", y = "Number of Apperances on the Top 100") +
    theme(plot.title = element_text(size = 15, hjust = -.3, face = "bold"), axis.title = element_text(size = 12)) +
    geom_text(aes(label=number_apperance), hjust = 2, size = 3, color = 'white') +
    coord_flip()
```

So, It seems like **Ed Sheeran** and **The Chainsmokers** rocked the 2017 with 4 different songs on the Top list, while **Drake**  and **Martin Garrix** are following them with 3 different songs.
